---
title: "Homework 3"
author: "Ben Wollack"
date: "2026/05/28"
format:
  pdf:
    toc: true
editor: source
execute:
  message: false
  warning: false
---

# Set Up

```{r}
#| label: read-in

# read in packages
library(tidyverse)
library(here)
library(janitor)
library(readxl)

# read in data
kelp <- read_csv(here("data", "temp-kelp.csv"))
walk <- read_csv(here("data", "ben-data-tracker.csv"))
```

# Problem 1

## a.

To determine the strength of the relationship between temperature and kelp elongation rate, we will need a correlation test, of which there are two: Pearson's correlation (r) and Spearman's correlation ($\rho$). Both require independent observations. Pearson's correlation is the parametric version and also requires data that appears linear and continuous. Spearman's correlation is the nonparametric version and requires data that is monotonic, meaning that as one variable increases, the other variable, on average, strictly increases or decreases; the trend cannot go up and down. 

## b.

```{r}
#| label: kelp-visual

# create base layer using kelp data
# predictor (x-axis): temperature
# response (y-axis): kelp_elong
ggplot(data = kelp,
       aes(
         x = temp_c,
         y = kelp_elong
       )) +
  # add points, which creates a scatterplot for the data
  geom_point(
    # change color to a kelp-like color
    color = "seagreen4",
    # change default shape
    shape = 17
  ) +
  # change theme to remove gridlines
  theme_classic() +
  # rename x and y axes to make them more readable
  labs(
    x = "Temperature (celsius)",
    y = "Kelp Elongation (cm/day)"
  )
```

There may be a slight curve in the data, but a linear model also seems reasonable. We can check using diagnostic plots.

## c.

Since the data appears linear, we will run a Pearson's correlation, which requires checking for linearity, the normality of the residuals and the homoscedasticity of the residuals.

### Diagnostic Plots

```{r}
#| label: kelp-model

# create a linear model in an object called kelp_model
# this lets us later test for linearity
kelp_model <- lm(
  # formula: kelp elongation as a function of temperature
  kelp_elong ~ temp_c,
  # use kelp data
  data = kelp
)
```

```{r}
#| label: diagnostic-kelp

# create diagnostic plots for the kelp model, displayed in a 2x2 grid
par(mfrow = c(2, 2))
plot(kelp_model)
```

The QQ plot of residuals has a few hiccups, but the data is close to the reference line with no obvious pattern, so the residuals are normally distributed.

The residuals are also homoscedastic since neither the residual vs fitted nor the scale-location model show clear patterns.

As a bonus, the residuals vs leverage plot also shows no outliers since all points are inside Cook's Distance.

### Correlation Test

```{r}
#| label: correlation-test

# run pearson's correlation test
cor.test(kelp$temp_c, kelp$kelp_elong,
         method = "pearson")
```

To prepare for writing about the results, here are the important outputs.

t = -5.189

df = 30

p < 0.001

95% CI: [-0.836, -0.446]

r = -0.688

## d.

To evaluate the strength of the relationship between water temperature (in Celsius) and kelp elongation rate (in cm per day), we used Pearson's correlation. We chose this because the scatterplot of data showing this relationship appeared linear, and the residuals were normally distributed and showed homoscedasticity.

We found a moderate, inverse relationship between water temperature and kelp elongation rate (Pearson's r = -0.688, t(30) = -5.189, p < 0.001, $\alpha$ = 0.05).

## e.

We found a moderate, inverse relationship between water temperature and kelp elongation rate (Pearson's r = -0.688, t(30) = -5.189, p < 0.001, $\alpha$ = 0.05). Our model predicts that as water temperature increases, kelp will grow more slowly. As ocean temperatures warm due to climate change, kelp growth rate will likely decrease and threaten the health of kelp forest ecosystems.

## f.

The other test that could have been run was a Spearman's correlation, the non parametric variant of Pearson's correlation. The relationship appears monotonic, so the assumptions for this test were met too.

```{r}
#| label: kelp-spearman

# run spearman's correlation test
# without the exact = false, the test returns an error since some data is of equal value and thus the same rank
cor.test(kelp$temp_c, kelp$kelp_elong,
         method = "spearman",
         exact = FALSE)
```

After accounting for ties, the test returns a similar p-value (< 0.001) and correlation ($\rho$ = -0.689). Therefore, we will interpret this test similarly. 

We found a moderate, inverse relationship between water temperature and kelp elongation rate (Spearman's $\rho$ = -0.689, S = 9216.1, p < 0.001, $\alpha$ = 0.05). Our model predicts that as water temperature increases, kelp will grow more slowly. As ocean temperatures warm due to climate change, kelp growth rate will likely decrease and threaten the health of kelp forest ecosystems.

# Problem 2

## a.

```{r}
#| label: plot-one

# create a new object to clean data
walk_clean <- walk |> 
  # remove uppercases and add underscores
  clean_names() |> 
  # remove N/A values
  drop_na()
# create a jitterplot for the data using walk_clean
ggplot(walk_clean,
       # x-axis: lunch location
       # y-axis: distance walked (in km)
       # color: set color by lunch location
       mapping = aes(x = lunch_location,
                     y = distance_walked_km,
                     color = lunch_location)) +
  # I want my y-axis to start at 0 so that scaling is visually accurate
  # Code from a geeksforgeeks forum
  scale_y_continuous(limits=c(0, 14)) +
  # Create a jitterplot
  geom_jitter(
    # stop vertical jitter
    height = 0,
    # narrow horizontal jitter
    width = 0.2) +
  # setting theme to black-white
  theme_bw() +
  # removing legend
  theme(legend.position = "None") +
  # renaming axes and title
  labs(
    # renaming x-axis
    x = "Lunch Location",
    # renaning y-axis
    y = "Distance Walked (km)",
    # renaming title
    title = "Distance Walked by Lunch Location",
    # subtitle for including last observation
    subtitle = "Last Observation: May 20th"
  ) +
  # changing colors of various elements to improve contrast
  theme(
    # change x-axis color
    axis.text.x = element_text(color = "black"),
    # change y-axis color
    axis.text.y = element_text(color = "black"),
    # change panel background
    panel.background = element_rect(fill = "lemonchiffon"),
    # change panel lines to blend in with background
    panel.grid = element_line(color = "lemonchiffon")
    ) +
  # change colors of data points
  scale_colour_manual(values = c(
    # change campus to a gold color to match UCSB :)
    "Campus" = "goldenrod4",
    # change IV to a lovely green
    "IV" = "green4",
    # change Portola to red
    "Portola" = "tomato3",
    # change prepped to a blue color
    "Prepped" = "cornflowerblue"
  ))
```

```{r}
#| label: plot-two

# create a scatterplot for the data using walk_clean
ggplot(walk_clean,
       # x-axis: high temperature (in celsius)
       # y-axis: distance walked (in km)
       # color: set color by lunch location
       mapping = aes(x = high_temperature_c,
                     y = distance_walked_km)) +
  # once again starting y from 0
  # It would not make sense to set temperature to 0 too
  # (this would make it hard to visualize the data's spread), 
  # so I am leaving the x-axis as is
  scale_y_continuous(limits=c(0, 14)) +
  # create a scatterplot with pink points
  geom_point(color = "violetred4") +
  # setting theme to classic
  theme_classic() +
  # removing legend
  theme(legend.position = "None") +
  # renaming axes and title
  labs(
    # renaming x-axis
    x = "High Temperature (celsius)",
    # renaning y-axis
    y = "Distance Walked (km)",
    # renaming title
    title = "Distance Walked by High Temperature"
  ) +
  # changing colors of various elements to improve contrast
  theme(
    # change x-axis color
    axis.text.x = element_text(color = "black"),
    # change y-axis color
    axis.text.y = element_text(color = "black"),
    # change panel background
    panel.background = element_rect(fill = "cornsilk4"),
    # change panel line colors to remove them
    panel.grid = element_line(color = "cornsilk4"))
```

## b.

# Problem 3

## a.

## b.

## c.

## d.

## e.

# Problem 4

## a.

## b.

## c.

## d.
